Event History Analyses for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-23 manuscript/manuscript.tex
update ms. and notes

In this paper, we focus on a distributional method for time-to-event data known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before (Allison, 1982, 2010; Singer \& Willett, 2003), and our primary aim here is to introduce a set of tutorials, which explain \textbf{how} to do such analyses, rather than repeat in any detail \textbf{why} you should do them.

@@

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

In summary, even though event history analyses is a widely used statistical tool and there already exist many excellent reviews (REFs) and tutorials (Allison, 2010) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps.
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using hazard analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time hazard analysis on time-to-event data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

\section{2. A brief introduction to hazard analysis}\label{a-brief-introduction-to-hazard-analysis}

For a comprehensive background context to hazard analysis, we recommend several excellent textbooks (Singer \& Willett, 2003). Likewise, for general introduction to understanding regression equations, we recommend several introductory level textbooks (REFs). Our focus here is not on providing a detailed account of the underlying regression equations, since this topics has been comprehensively covered many times before. Instead, we want to provide an intuition to how EHA works in general as well as in the context of experimental psychology. As such, we only supply regression equations in the supplementary material (part D) and then refer to them in the text whenever relevant.

\subsection{2.1 Basic features of hazard analysis}\label{basic-features-of-hazard-analysis}

To apply event history analysis (EHA), one must be able to:

@@

Discrete-time EHA focuses on the discrete-time hazard function and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard probability gives you the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500{]}, then only 3\% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.

@@

\subsection{2.2 Hazard analysis in the context of experimental psychology}\label{hazard-analysis-in-the-context-of-experimental-psychology}

@@

In the context of experimental psychology, it is common for participants to be presented with a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time hazard analysis can be extended with a discrete-time SAT analyis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for early responses (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for later bins, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

\begin{itemize}
\tightlist
\item
  1a. Wrangle raw data and descriptive stats for one independent variable.
\item
  1b. Wrangle raw data and descriptive stats for two independent variables.
\item
  2a. Bayesian multilevel modeling for h(t)
\item
  2b. Bayesian multilevel modeling for ca(t)
\item
  3a. Frequentist multilevel modeling for h(t)
\item
  3b. Frequentist multilevel modeling for ca(t)
\end{itemize}

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions about the timecourse of psychological processes. Second, we want to produce two different data sets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as `person-trial' data and `person-trial-bin' data. The `person-trial' data (Table 1) will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. This data set is used when fitting conditional accuracy models.

@@

In contrast, the `person-trial-bin' data (Table 2) has a different, more extended structure, which indicates in which bin a response occurred, if at all, in each trial. Therefore, the `person-trial-bin' dataset generates a 0 in each bin until an event occurs and then it generates a 1 to signal an event has occurred in that bin. This data set is used when fitting hazard models. It is worth pointing out that there is no requirement for an event to occur at all (in any bin), as maybe there was no response on that trial or the event occurred after the time window of interest. Likewise, when the event occurs in bin 1 there would only be one row of data for that trial in the person-trial-bin data set.

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t), ca(t), and P(t) -- see part A of the supplementary material. To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our custom function ``censor()'', as follows.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias (REFs). The person-trial-bin oriented dataset has one row for each time bin of each trial that is at risk for event occurrence. The variable ``event'' in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.

The next step is to plot the data using our custom function plot\_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant. It is important to visually inspect the functions first for each participant, in order to identify possible cheaters (e.g., a flat conditional accuracy function at .5 indicates (s)he was only guessing), outlying individuals, and/or different groups with qualitatively different behavior.

Table 3 shows the life table for condition ``blank'' (no prime stimulus presented) for participant 6 -- compare to Figure 1. A life table includes for each time bin, the risk set (number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

Figure 4 displays the discrete-time hazard, survivor, conditional accuracy, and probability mass functions for each prime condition for participant 6. By using discrete-time h(t) functions of event occurrence -- in combination with ca(t) functions for two-choice tasks -- one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any data set.

@@

\caption{Sample-based discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of prime type (blank, congruent, incongruent) and mask type (no mask, relevant, irrelevant, lines).}\label{fig:plot-2ivs}
\end{figure}

@@

In this third tutorial, we illustrate how to fit Bayesian hazard regression models to the masked response priming data set used in Tutorial 1a. Fitting (Bayesian or non-Bayesian) regression models to the data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below). We provide relevant example regression formulas in part D of the supplementary material.

@@

\NormalTok{model\_M1 }\OtherTok{\textless{}{-}}
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ M1\_data,}
       \AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
\NormalTok{       event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{+}
\NormalTok{                           (d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+}\NormalTok{ condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors\_M1,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }\AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"Tutorial\_2\_Bayesian/models/model\_M1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

So far, we assumed that the effect of our predictors condition and centered trial number are the same in each time bin. However, the descriptive plots suggest that the effect of condition can vary across time bins. Previous research has indeed shown that psychological effects typically change over time (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009).
For the third model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables prime-target congruency (variable ``condition'') and centered trial number (variable ``trial\_c'').

@@

Based on Figure 6 and Table 4, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240{]} on cloglog-hazard, relative to the cloglog-hzard estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials, in bin (200,240{]} of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360{]}, (360,400{]}, and (400,440{]}. For example, in bin (320,360{]}, the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and target-triggered response (e.g., right)

@@

Panis and Schmidt (2016) distinguished between automatic response competition effects (bottom-up lateral inhibition between response channels), active and global inhibition effects (top-down nonselective response inhibition), and active and selective inhibition (top-down selective response inhibition). While automatic response competition can be expected to be present in the incongruent trials throughout the experiment, active and global inhibition effects might be present in both congruent and incongruent (unmasked) prime trials. In other words, people learn that the prime-triggered response is premature and that they have to temporarily slow down (increase the global response threshold) in order to allow gating of the target stimulus. This global inhibitory effect might become visible in the congruent (compared to no-prime) trials towards the end of the experiment, while it might be masked by the automatic inhibitory effect of response competition in the incongruent trials.
Interestingly, while Panis and Schmidt (2016) did not test interactions between congruency and trial number, they concluded that active (i.e., top-down) response inhibition starts around 360 ms after the onset of the second stimulus (the target stimulus in no-mask trials), which coincides with the onset of the negative effect of congruent primes observed here in trial 1500.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. For illustration purposes, we only fitted the effects of model M4 (see Tutorial 2a).

To make inferences from the parameter estimates in model M4\_ca, we summarize the draws from the posterior distributions of the effect of congruent and incongruent primes relative to the blank prime condition, in each time bin for trial numbers 500, 1000, and 1500, in terms of point and interval estimates.

@@

In this fifth tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS. The resulting hazard model is called M3\_f.

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331) when one is interested in studying the detailed shape of a RT distribution (see also Figure 1 in Panis, Schmidt, et al., 2020). Therefore, when the goal is to study how psychological effects change over time, hazard functions are preferred.

>>>>

In this paper, we focus on a distributional method for time-to-event data known as \emph{discrete-time event history analysis}, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of event history analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before (Allison, 1982, 2010; Singer \& Willett, 2003), and our primary aim here is to introduce a set of tutorials, which explain \textbf{how} to do such analyses, rather than repeat in any detail \textbf{why} you should do them.

@@

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the event history analysis of RT data for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of multilevel regression models for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to event history analysis. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

In summary, even though event history analysis is a widely used statistical tool and there already exist many excellent reviews (REFs) and tutorials (Allison, 2010) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps.
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using event history analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time event history analysis on time-to-event data such as RT data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

\section{2. A brief introduction to event history analysis}\label{a-brief-introduction-to-event-history-analysis}

For a comprehensive background context to event history analysis, we recommend several excellent textbooks (Singer \& Willett, 2003). Likewise, for general introduction to understanding regression equations, we recommend several introductory level textbooks (REFs). Our focus here is not on providing a detailed account of the underlying regression equations, since this topics has been comprehensively covered many times before. Instead, we want to provide an intuition to how event history analysis works in general as well as in the context of experimental psychology. As such, we only supply regression equations in the supplementary material (part D) and then refer to them in the text whenever relevant.

\subsection{2.1 Basic features of event history analysis}\label{basic-features-of-event-history-analysis}

To apply event history analysis (EHA), a.k.a. hazard analysis, one must be able to:

@@

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard function gives you for each bin the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

{[}{[}this paragraph is not so clear yet\ldots{} Perhaps add error bars to the functions in Figure 2?{]}{]}
The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500{]}, then only 3\% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.

@@

\subsection{2.2 Event history analysis in the context of experimental psychology}\label{event-history-analysis-in-the-context-of-experimental-psychology}

@@

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task, i.e., a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA can be extended with a discrete-time SAT analysis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977). We refer to this extended analysis for choice RT data as EHA/SAT.

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for short waiting times (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for longer waiting times, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

\begin{itemize}
\tightlist
\item
  1a. Wrangle raw data and calculate descriptive stats for one independent variable.
\item
  1b. Wrangle raw data and calculate descriptive stats for two independent variables.
\item
  2a. Bayesian multilevel modeling for h(t)
\item
  2b. Bayesian multilevel modeling for ca(t)
\item
  3a. Frequentist multilevel modeling for h(t)
\item
  3b. Frequentist multilevel modeling for ca(t)
\end{itemize}

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions about the time course of psychological processes. Second, we want to produce two different data sets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as `person-trial' data and `person-trial-bin' data. The `person-trial' data (Table 1) will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. This data set is used when fitting conditional accuracy models (Tutorials 2b and 3b).

@@

In contrast, the `person-trial-bin' data (Table 2) has a different, more extended structure, which indicates in which bin a response occurred, if at all, in each trial. Therefore, the `person-trial-bin' dataset generates a 0 in each bin until an event occurs and then it generates a 1 to signal an event has occurred in that bin. This data set is used when fitting hazard models (Tutorials 2a and 3a). It is worth pointing out that there is no requirement for an event to occur at all (in any bin), as maybe there was no response on that trial or the event occurred after the time window of interest. Likewise, when the event occurs in bin 1 there would only be one row of data for that trial in the person-trial-bin data set.

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t), ca(t), and P(t) -- see part A of the supplementary material for their definitions. To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our custom function ``censor()'', as follows.

@@

Note that the censoring time should be a multiple of the bin width (both in ms). The censoring time should be a time point after which no informative responses are expected anymore. In experiments that implement a response deadline in each trial the censoring time can equal that deadline time point. Trials with a RT larger than the censoring time, or trials in which no response is emitted during the data collection period, are treated as right-censored observations in EHA. In other words, these trials are not discarded, because they contain the information that the event did not occur before the censoring time. Removing such trials before calculating the mean event time can introduce a sampling bias (REFs).

The person-trial-bin oriented dataset is created by our custom function ptb(), and it has one row for each time bin of each trial that is at risk for event occurrence. The variable ``event'' in the person-trial-bin oriented data set indicates whether a response occurs (1) or not (0) for each bin.

The next step is to set up the life table -- custom function setup\_lt() --, calculate the conditional accuracies -- custom function calc\_ca() --, and then plot the descriptive statistics using our custom function plot\_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant.

In general, it is important to visually inspect the functions first for each participant, in order to identify possible cheaters (e.g., a flat conditional accuracy function at .5 indicates (s)he was only guessing), outlying individuals, and/or different groups with qualitatively different behavior.

Table 3 shows the life table for condition ``blank'' (no prime stimulus presented) for participant 6. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), P(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

Figure 4 displays the discrete-time hazard, survivor, conditional accuracy, and probability mass functions for each prime condition for participant 6. By using discrete-time hazard functions of event occurrence -- in combination with conditional accuracy functions for two-choice tasks -- one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any data set.

@@

\caption{Sample-based discrete-time hazard, survivor, conditional accuracy, and probability mass functions for participant 6, as a function of prime type (blank, congruent, incongruent) and mask type (no mask, relevant, irrelevant, lines).}\label{fig:plot-2ivs}
\end{figure}

@@

In this third tutorial, we illustrate how to fit Bayesian hazard regression models to the RT data of the masked response priming data set used in Tutorial 1a. Fitting (Bayesian or non-Bayesian) regression models to time-to-event data is important when you want to study how the shape of the hazard function depends on various predictors (Singer \& Willett, 2003).

@@

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable (event occurrence) is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below). We provide relevant example regression formulas in part D of the supplementary material.

@@

\NormalTok{model\_M1 }\OtherTok{\textless{}{-}}
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ M1\_data,}
       \AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
\NormalTok{       event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{+}
\NormalTok{                           (d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{                           condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors\_M1,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }\AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"Tutorial\_2\_Bayesian/models/model\_M1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

So far, we assumed that the effect of our predictors condition and centered trial number are the same in each time bin. However, the descriptive plots suggest that the effect of prime-target congruency varies across time bins. Previous research has shown that psychological effects typically change over time (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009).
For the third model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and relax the proportionality assumption for the predictor variables prime-target congruency (variable ``condition'') and centered trial number (variable ``trial\_c'').

@@

Based on Figure 6 and Table 4, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240{]} on cloglog-hazard, relative to the cloglog-hzard estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials in bin (200,240{]} of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360{]}, (360,400{]}, and (400,440{]}. For example, in bin (320,360{]}, the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and the target-triggered response (e.g., right)

@@

Panis and Schmidt (2016) distinguished between automatic response competition effects (bottom-up lateral inhibition between response channels), active and global inhibition effects (top-down nonselective response inhibition), and active and selective inhibition (top-down selective response inhibition). While automatic response competition can be expected to be present in the incongruent trials throughout the experiment, active and global response inhibition effects might be present in both congruent and incongruent (unmasked) prime trials. In other words, people learn that the prime-triggered response is premature and that they have to temporarily slow down (increase the global response threshold) in order to allow gating of the response to the target stimulus. This global inhibitory effect becomes visible in the congruent (compared to no-prime) trials towards the end of the experiment, while it might be masked by the automatic inhibitory effect of response competition in the incongruent trials.
Interestingly, while Panis and Schmidt (2016) did not test interactions between congruency and trial number, they concluded that active (i.e., top-down) response inhibition starts around 360 ms after the onset of the second stimulus (the target stimulus in no-mask trials), which coincides with the onset of the negative effect of congruent primes observed here in trial 1500.

@@

In this fourth tutorial, we illustrate how to fit a Bayesian regression model to the timed accuracy data from the masked response priming data set used in Tutorial 1a. The general process is similar to Tutorial 2a, except that (a) we use the person-trial data set, (b) we use the logit link function, and (c) we change the priors. For illustration purposes, we only fitted the effects of model M4 (see Tutorial 2a) in the conditional accuracy model called M4\_ca.

To make inferences from the parameter estimates in model M4\_ca, we summarize the draws from the posterior distributions of the effects of congruent and incongruent primes relative to the blank prime condition, in each time bin for trial numbers 500, 1000, and 1500, in terms of point and interval estimates.

@@

In this fifth tutorial we illustrate how to fit a multilevel hazard regression model in the frequentist framework, for the data set used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set. For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS (REF). The resulting hazard model is called M3\_f.

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331) when one is interested in studying the detailed shape of a RT distribution (see also Figure 1 in Panis, Schmidt, et al., 2020). Therefore, when the goal is to study how psychological effects change over time, hazard and conditional accuracy functions are preferred.
